---
title: "basic-usage"
author: "Dekang Lv"
date: "May 2025"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic-usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Introduction
Thank you for your interest in the *nichetag* approach. This vignette describes how to apply the package to your intercellular RNA-barcode delivery experiment(Multiplexed Oligonucleotide Spatial Ancestry Interaction Coding, MOSAIC). MOSAIC utilizes conditionally expressed, proximally
transmissible sLP-mCherry-MCP-MS2-CellTag barcodes to trace the spatiotemporal lineages and communicating networks within intact tissues at single-cell resolution. The goal of the *nichetag* package is to use CellTag labels from single-cell transcriptomics data to reconstruct the lineage and spatial proximity relationships of mCherry-positive cells. To learn more about this type of experiments and details about the nichetag approach, please consult [Bin et al, 2025](insert paper URL here). This vignette aims at explaining how to apply a basic *nichetag* analysis.
    
This vignette will cover:

1. Package installation
2. Setting up your input data
3. Running the nichetag computational pipeline
4. Conclusion and information about the R session

# Package installation
You can install the package from github using
```{r, installation, eval=FALSE}
#Install dependencies
install.packages(c("Seurat","Polychrome","stringdist","stringr","ggplot2","cowplot","igraph","openxlsx"))
if(!require(devtools)){
	install.packages("devtools")
}
#install nichetag
if(!require(nichetag)){
  devtools::install_github("woshijiaomu/nichetag")
}
```

# Setting up your input data

Now that the package is installed, we load the package and attach its example data. We also load Seurat that will be used to manipulate Seurat Objects in R.

```{r, load_package, eval = TRUE}
library(nichetag)
data(example)
```

The example data is a single-cell transcriptome (10XGenomics) which was performed in the mCherry+ cells sorted from early (2-week) metastatic lung tissues 4 days after Dox exposure. In addition to gene expression data, cell classification information and CellTag expression profiles within the object are utilized to identify spatial proximity relationships. In this vignette we will use a temporary directory, but you can use the directory of your choice. That is all we need in terms of inputs! Below is the pre-processsing workflow for obtaining this data:

1.To enable the detection of exogenous genes and tags, a custom mouse reference genome was generated for Cell Ranger (version 8.0.1). The coding sequences of mCherry, BFP, and all relevant CellTag sequences were appended to the mouse genome FASTA file, and their corresponding gene annotations were added to the GTF file under a new gene category. Using these modified files, a new Cell Ranger reference was built using the “cellranger mkref” command. Raw data were then processed using the cellranger count pipeline, aligning reads to the custom reference genome and generating filtered gene-
barcode matrices for downstream analysis.
2.The matrix data was analyzed using the Seurat package (version 4.0.0), and cell clusters were annotated using the SingleR package (version 2.10.0).

# Running the nichetag computational pipeline

Now that we have our gene expresion data, celltag expression profiles and cell type annotation, we have everything we need to run the pipeline.

All the pipeline is mainly packaged into a single function, `Dnichenetwork()`.

Here is a description of the basic arguments it requires:

1. scObject: Seurat object, input seurat object in which tag expression matrix is merged in gene expression matrix
2. groupby: character, a column name in meta data of the Seurat object, used for clone definition
3. share_method: min,max,mean, method to calculate connection strength between different clones
4. direction: True or False, whether or not to distinguish tag-sender or tag-receiver

We have everything we need in our input object to fill these arguments:
```{r, inspect_input_scObject}
#calculate connectome
nnt=Dnichenetwork(scObject,groupby="cell_clusters")
summary(nnt)
```

This function returns a list containing connectome information for all FCMs, which mainly includes:

1.tag_matrix: data.frame, CellTag expression matrix of all cells
2.cell_type: factor, cell type or state used for clone defination
3.cell_tags: numeric, total expression counts of each CellTag
4.niche_size: integer, covered cell number of each CellTag
5.niche_celltypes: integer, covered cell types or cell states by each CellTag
6.niche_interaction: data.frame, overlapped cell numbers of FCM-FCM pairs
7.cell_barcode: character, code of each cell, celltagcode+celltypecode
8.clone list: each element represent a CellTag expression matrix of each clone
9.clone_size: integer, covered cell number of each clone
10.niche list: mean CellTag expression of covered cells of all covered clone for each CellTag
11.clone_interaction: data.frame, overlapped tag expression of clone-clone pairs
12.cloneID: integer, order of clones when make graph with graph_from_data_frame in igraph

Visualization for connectome data encompass the following functions:

1. print_nichenetwork: Clone network - each vertex represents a clone, and edges between vertices indicate shared CellTags between two clones
2. print_nichetag: Bar plot - displays the cell number and cell types covered by each FCM
3. print_clustertag: Density plot - shows the distribution of CellTag expression levels within each cell cluster
4. tag_cancer_noncancer: Bar plot - comparatively displays the expression levels of each CellTag in cancer cells and stromal cells
5. tag_cci: Bar plot - shows the distribution of sender and receiver cell counts corresponding to different CellTags
6. clonetype: Bar plot - displays the cell types and quantities of clones within each FCM
7. tag_cellclonetype: Bar plot - shows the types and quantities of clones or cells covered by each CellTag
8. print_nichenet: Clone network - each vertex represents an FCM, and edges between vertices indicate the presence of shared CellTags between two FCMs

We run these functions for visualization:
```{r, visualization}
#draw clone-clone network
print_nichenetwork(nnt,file="nichenetwork.pdf",vertex.label.cex=0.1,vsize=3,esize=1,margin=c(0,0.3,0,0))
#draw quanlity control figures
print_nichetag(nnt, file = "nichetag.pdf")
print_clustertag(nnt, file = "cluster.pdf")
tag_cancer_noncancer(nnt, file = "celltype.pdf")
tag_cci(nnt, file = "cci.pdf")
clonetype(nnt, file="clonetype.pdf")
tag_cellclonetype(nnt, file="tag2celltype_clonetype.pdf")
#draw niche-niche network
print_nichenet(nnt,file="niche2nichenetwork.pdf",vsize = 10)
```

Export Data

In addition to the visualization parts above, we can also extract from the connectome object: 
1. the list of clones contained in each FCM
2. the list of cells contained in each clone
3. the average expression matrix for each clone

The information can be written to local CSV files, which can be useful to further analyze the data in R:
```{r, export_data}
#clones and the cells it contains
clone2cell=nnt$clone
clone2cell.vec=unlist(lapply(names(clone2cell),function(x){
  y=clone2cell[[x]]
  res=rep(x,nrow(y))
  names(res)=rownames(y)
  res
}))
clone2cell.df=data.frame(clone_code=clone2cell.vec,cell=names(clone2cell.vec))
clone2cell.df$clone_ID=nnt$cloneID[clone2cell.df$clone_code]
write.csv(clone2cell.df,file = "clone2cell.csv")

#FCMs and the clones it contains
niche2clone=nnt$niche
niche2clone.vec=unlist(lapply(names(niche2clone),function(x){
  res=paste(x,names(niche2clone[[x]]),sep=":")
  res
}))
niche2clone.df=as.data.frame(stringr::str_split_fixed(niche2clone.vec,":",n=2))
colnames(niche2clone.df) = c("niche","clone_code")
niche2clone.df$clone_ID=nnt$cloneID[niche2clone.df$clone_code]
write.csv(niche2clone.df,file = "niche2clone.csv")

#clone expression matrix
clonematrix=Clone_expr(scObject,nnt,seurat_layer="counts")
write.csv(clonematrix,"clone_matrix.csv")
```

# Conclusion

Thank you for following this vignette, I hope you made it through the end without too much headache and that it was informative. If you encounter bugs, feel free to raise an issue on nichetag's [github](https://github.com/woshijiaomu/nichetag/issues).

# Information about the R session when this vignette was built
```{r, sessionInfo}
sessionInfo()
```
